Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-09-28 Dail_Mad_ext.tex
Bayes

Maximum likelihood estimation of parameters in models with random
effects requires removing the random effects by integration or
summation. For Markovian models, this can be done recursively, making
the likelihood tractible. A general expression for the likelihood of
the DM model is:
\begin{equation}
  \mathcal{L}(\Lambda, \Phi, \mathbf{p} | \{n_{it}\}) =
  \prod_{i=1}^R \left\{ \sum_{N_{i1}=0}^{\infty}
  [n_{i1}|N_{i1}][N_{i1}|\Lambda] \left\{ \prod_{t=T}^2 \left\{ \sum_{N_{it}=0}^{\infty}
  [n_{it}|N_{it}][N_{it}|N_{it-1}] \right\} \right\} \right\}
  \label{eq:like}
\end{equation}

@@

Bayesian inference has several appealing features. First, it allows
direct probability statements to be made about a hypothesis given
data. Second, priors are cool. We can combine data from multiple
sources or use existing estimates of parameters as prior
distributions. MCMC has a bad rap for being slow, but actually, in the
case of state-space models it may be one of the few options for
estimation, and in fact may be faster than maximizing likelihoods.

Here is some BUGS code:

\begin{verbatim}
model {
lambda ~ dunif(0, 5)
r ~ dunif(-5, 5)
p ~ dunif(0,1)
for(i in 1:nSites) {
  N[i,1] ~ dpois(lambda)
  y[i,1] ~ dbin(p, N[i,1])
  for(t in 2:nYears) {
    S[i,t-1] ~ dbin(omega, N[i,t-1])
    G[i,t-1] ~ dpois(gamma)
    N[i,t] <- S[i,t-1] + G[i,t-1]
    n[i,t] ~ dbin(p, N[i,1])
    }
  }
}
\end{verbatim}

>>>>

Maximum likelihood estimation of parameters in models with random
effects requires removing the random effects by integration or
summation. For Markovian models, this can be done recursively, making
the likelihood tractible, the general expression of which can be written:
\begin{equation}
  \mathcal{L}(\Lambda, \Phi, \mathbf{p} | \{n_{it}\}) =
  \prod_{i=1}^R \left\{ \sum_{N_{i1}=0}^{\infty}
  [n_{i1}|N_{i1}][N_{i1}|\Lambda] \left\{ \prod_{t=T}^2 \left\{ \sum_{N_{it}=0}^{\infty}
  [n_{it}|N_{it}][N_{it}|N_{it-1}] \right\} \right\} \right\}
  \label{eq:like}
\end{equation}
This likelihood can be maximized numerically using the \textbf{R}
package \texttt{unmarked} \citep{fiske_chandler:2011}. Examples are
given in the Appendix.

@@

Bayesian inference has several appealing features. First, it allows
direct probability statements to be made about a hypothesis given
data. [EXPLAIN POSTERIORS AND PRIORS]
Second, Bayesian methods offer straight-forward appraoches for
combine data from multiple
sources or use existing estimates of parameters as prior
distributions. In practice, simulation methods such as Markov chain
Monte Carlo (MCMC) are used to simulate the posterior
distributions. Although MCMC can be slow, our experience is that it
may actually converge faster than the amount of time required to
maximize the likelihood shown in Eq.\ref{eq:like}, especially when the
upper bound of the summation must be high, e.g. $K>200$. Furthermore,
MCMC may be the only possible option for estimating parameters using
some of the extensions described below, such as allowing for
additional random effects.

Writeing custom MCMC algorthims is often tedious and foreign to
ecologists, but software packages such as \textbf{WinBUGS} and
\textbf{JAGS} overcome the technical difficulties by allowing users to
specify the model using a simple symbolic descriptions. For example,
here is \textbf{BUGS} code for the ``constant'' model described above.

\begin{verbatim}
model {
lambda ~ dunif(0, 5)
omega ~ dunif(0,1)
gamma ~ dunif(0, 10)
p ~ dunif(0,1)
for(i in 1:nSites) {
  N[i,1] ~ dpois(lambda)
  y[i,1] ~ dbin(p, N[i,1])
  for(t in 2:nYears) {
    S[i,t-1] ~ dbin(omega, N[i,t-1])
    G[i,t-1] ~ dpois(gamma)
    N[i,t] <- S[i,t-1] + G[i,t-1]
    n[i,t] ~ dbin(p, N[i,1])
    }
  }
}
\end{verbatim}
